\Ace: An Actively-Typed Compilation Environment\\for High-Performance Computing

### 2012-09-20 ipdps13/ace-ipdps13.tex
Cleaned up abstract.

\begin{abstract}
In this paper, we introduce the Ace compilation environment and demonstrate its practicality as a foundational tool for both high-performance computing research and in practice. 
Ace consists of a statically-typed language that shares syntax with Python and uses the Python language itself as a multi-purpose compile-time metalanguage. Unlike standard languages, however, the Ace compiler consists only of a minimal core that delegates control over most aspects of front-end compilation to {\em active libraries} using a novel extension mechanism that we call {\em active type-checking and translation (\ATT)}. Users specify new primitive types and operations by equipping type definitions, which are first-class objects in the metalanguage, with designated methods that the compiler selectively invokes when type checking and translating expressions. Using \ATT, we implement the primitives of the C99 and OpenCL languages as modules within Ace. We then give examples of several first-class parallel abstractions that can be used alongside these basic primitives, demonstrating the foundational role that an actively-typed system like Ace can play for researchers developing language-based programming abstractions.

Ace also supports more conventional forms of structured and unstructured metaprogramming and Ace programs can be compiled and launched from Python directly. We used these features, along with the OpenCL language module, to design a scientific simulation framework that allows users to modularly specify and orchestrate the execution of parameterized families of simulations on clusters of GPUs. This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing  evidence that Ace is also useful in practice today.
\end{abstract}

>>>>

\begin{abstract}
In this paper, we introduce the Ace compilation environment and demonstrate its practicality as a foundational tool for both high-performance computing research and in practice. 
Ace consists of a statically-typed language that shares syntax with Python embedded within the Python language itself, which serves as the compile-time metalanguage. Unlike standard languages, Ace consists only of a minimal core that delegates control over most aspects of front-end compilation to {\em active libraries} using a novel extension mechanism that we call {\em active type-checking and translation (\ATT)}. Users specify new primitive types and operations by equipping type definitions, which are first-class objects in the metalanguage, with  methods that the compiler selectively invokes when type checking and translating expressions. \ATT~ allows us to  implement the primitives of the C99 and OpenCL languages as simple modules within Ace. We then give examples of several first-class parallel abstractions that can be defined and used alongside these basic primitives, demonstrating the foundational role that an actively-typed system like Ace can play for researchers and domain experts developing first-class abstractions as well as practitioners wishing to combine multiple such abstractions without interfacing between many distinct languages.

Active library support in Ace also allows for conventional forms of structured and unstructured metaprogramming, and programs can be compiled and launched from Python directly. We used these features, along with the OpenCL language module and some extensions, to design a scientific simulation framework that allows users to modularly specify, compile and orchestrate the execution of parameterized families of scientific simulations on clusters of GPUs. This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing initial evidence that Ace can be used at scale in practice today.
\end{abstract}
